Multi-level generalizations of collective spin dynamics and squeezing

### 2019-09-27 collective_sun/sun_notes.tex
added discussion mapping drive operators onto spin operators of a collective SU(2) system

The GGM operators in \eqref{eq:GGM_diag} and \eqref{eq:GGM_off_diag}
provide a convenient operator basis to describe dynamics obeying
SU($n$) symmetry.  An external driving field addressing $n$-level
nuclear spins, however, will generally violate SU($n$) symmetry, and
instead obeys the symmetries of the {\it polarization} or {\it
  transition operators} defined by\cite{kryszewski2006alternative,
  bertlmann2008bloch}
\begin{align}
  T_{LM} \equiv \sqrt{\f{2L+1}{2I+1}} \sum_{\mu,\nu}
  \bk{I\mu;I\nu,LM} \op{\mu}{\nu},
  \label{eq:trans_ops}
\end{align}
where $L\in\set{0,1,\cdots,n-1}$ and $M\in\set{-L,-L+1,L}$ index a
total spin and its projection onto a quantization axis;
$I\equiv\p{n-1}/2$ is the maximal angular momentum of an $n$-level
spin; $\mu,\nu\in\set{-I,-I+1,\cdots,I}$ index projections of an
$n$-level nuclear spin onto a quantization axis; and
$\bk{I\mu;I\nu,LM}$ is a Clebsch-Gordan coefficient.  Similarly to the
GGM operators in \eqref{eq:GGM_diag} and \eqref{eq:GGM_off_diag}, the
transition operators in \eqref{eq:trans_ops} provide a complete basis
for the space of operators on the Hilbert space of an $n$-level
system, and satisfy the orthonormality condition
\begin{align}
  \obk{T_{LM}|T_{L'M'}} = \delta_{LL'} \delta_{MM'},
\end{align}
which implies
\begin{align}
  H_{\t{int}} = -\f12 u \sum_{p,q} {\v T^{(p)}}^\dag \c \v T^{(q)}
  = -\f12 u \v\T^\dag \c \v\T,
  &&
  \v\T \equiv \v\T\equiv\sum_p\v T^{(p)},
  \label{eq:H_int_trans}
\end{align}
where $\v T$ is a vector of all transition operators in
\eqref{eq:trans_ops}.

@@

Unlike the GGM operators in \eqref{eq:GGM_diag} and
\eqref{eq:GGM_off_diag}, the transition operators in
\eqref{eq:trans_ops} are not self-adjoint.  In order to express
Hamiltonians in terms of self-adjoint operators, we define the {\it
  drive operators}
\begin{align}
  D_{L,0} \equiv \sqrt{2}~ T_{L,0},
  &&
  D_{LM} \stackrel{M>0}{\equiv} -\p{T_{L\Delta} + T_{L\Delta}^\dag},
  &&
  D_{LM} \stackrel{M<0}{\equiv} i\p{T_{L\Delta} - T_{L\Delta}^\dag},
  \label{eq:drive_ops}
\end{align}
for integer $M$ with $\abs{M}\le L$.  Similarly to the GGM operators
in \eqref{eq:GGM_diag} and \eqref{eq:GGM_off_diag}, the drive
operators are equal to the Pauli operators in the case of $n=2$, and
satisfy the orthonormality condition
\begin{align}
  \obk{D_{LM}|D_{L'M'}} = 2\delta_{LL'}\delta_{MM'},
\end{align}
where $a,b$ index any drive operator in \eqref{eq:drive_ops}.  These
operators have the convenient feature that $D_{1,1}$ is (up to a gauge
transformation) proportional to the Hamiltonian induced on an
$n$-level spin by a classical driving laser.  In terms of the drive
operators,
\begin{align}
  H_{\t{int}} = -\f14 u \sum_{p,q} \v D^{(p)}\c\v D^{(q)}
  = -u \v\D \c \v\D,
  &&
  \v\D \equiv \f12 \sum_p \v D^{(p)},
  \label{eq:H_int_drive}
\end{align}
where $\v D$ is a vector of all drive operators in
\eqref{eq:drive_ops}.

@@

Here we take the single-particle spin-orbit coupled Hamiltonian
$H_{\t{lat}}^{(\phi)}$ in \eqref{eq:H_lat_SOC}, and derive its
expression in \eqref{eq:H_lat_drive} in terms of the drive operators
$D_{LM}$.  In the basis of ordinary spin operators $s_{\mu\mu}$, this
Hamiltonian takes the form
\begin{align}
  H_{\t{lat}}^{(\phi)}
  = -J \sum_{q,\mu} \cos\p{qa+\mu\phi} s_{\mu\mu}^{(q)}
  \label{eq:H_lat_spin}
\end{align}
Defining the diagonal ``spin-wave'' operators
\begin{align}
  \omega_{\phi,+} \equiv \sum_\mu \cos\p{\mu\phi} s_{\mu\mu},
  &&
  \omega_{\phi,-} \equiv \sum_\mu \sin\p{\mu\phi} s_{\mu\mu},
\end{align}
which are respectively even and odd under spin inversion
($\mu\to-\mu$), we can simplify
\begin{align}
  \sum_\mu \cos\p{qa+\mu\phi} s_{\mu\mu}
  = \cos\p{qa} \omega_{\phi,+} - \sin\p{qa} \omega_{\phi,-},
\end{align}
and resolve the identity (super-)operator $\I$ with
$\I \omega_{\phi,\pm} = \omega_{\phi,\pm}$ in the basis of the drive
operators, finding
\begin{align}
  \sum_\mu \cos\p{qa+\mu\phi} s_{\mu\mu}
  = \cos\p{qa} \f12 \sum_{L,M} \obk{\omega_{\phi,+}|D_{LM}} D_{LM}
  - \sin\p{qa} \f12 \sum_{L,M} \obk{\omega_{\phi,-}|D_{LM}} D_{LM}.
\end{align}
We now note that the drive operators $D_{LM}$ are strictly
off-diagonal for $M\ne0$, and furthermore that $D_{L,0}$ with even
(odd) $L$ is even (odd) under spin inversion, which implies
\begin{align}
  \sum_{L,M} \obk{\omega_{\phi,+}|D_{LM}} D_{LM}
  = \sum_{L\,\t{even}} \obk{\omega_{\phi,+}|D_{L,0}} D_{L,0},
\end{align}
and likewise with $\omega_{\phi,-}$ and odd $L$.  In total, we can
write
\begin{align}
  \sum_\mu \cos\p{qa+\mu\phi} s_{\mu\mu}
  = \cos\p{qa} \f12 \sum_{L\,\t{even}}
  \obk{\omega_{\phi,+}|D_{L,0}} D_{L,0}
  - \sin\p{qa} \f12 \sum_{L\,\t{odd}}
  \obk{\omega_{\phi,-}|D_{L,0}} D_{L,0},
\end{align}
or, more compactly,
\begin{align}
  \sum_\mu \cos\p{qa+\mu\phi} s_{\mu\mu}
  = \sum_L B_L\p{q,\phi} D_{L,0},
  \label{eq:compact_SOC_drive}
\end{align}
with effective driving field strengths
\begin{align}
  B_L\p{q,\phi}
  &\equiv \p{-1}^L f_L\p{qa} \f12
  \sum_\mu f_L\p{\mu\phi} \obk{s_{\mu\mu}|D_{L,0}} \\
  &= \p{-1}^L f_L\p{qa} \sqrt{\f{L+1/2}{n}}
  \sum_\mu \bk{I\mu;I\mu,L,0} f_L\p{\mu\phi},
\end{align}
where
\begin{align}
  f_L\p{\theta} \equiv
  \begin{cases}
    \cos\theta & L~\t{even} \\
    \sin\theta & L~\t{odd}
  \end{cases}.
\end{align}
Substituting the result in \eqref{eq:compact_SOC_drive} into the
single-particle Hamiltonian in \eqref{eq:H_lat_spin} yields
\begin{align}
  H_{\t{lat}}^{(\phi)} = -J \sum_{q,L} B_L\p{q,\phi} D_{L,0}^{(q)}.
\end{align}

>>>>

The GGM operators in \eqref{eq:GGM_diag} and \eqref{eq:GGM_off_diag}
provide a convenient operator basis to describe dynamics obeying
SU($n$) symmetry.  An external driving field addressing $n$-level
nuclear spins, however, will generally violate SU($n$) symmetry, and
instead obeys the symmetries of the {\it polarization} or {\it
  transition operators} defined by\cite{kryszewski2006alternative,
  bertlmann2008bloch}
\begin{align}
  T_{LM} \equiv \sqrt{\f{2L+1}{2I+1}} \sum_{\mu,\nu}
  \bk{I\mu;I\nu,LM} \op{\mu}{\nu},
  \label{eq:trans_ops}
\end{align}
where $L\in\set{0,1,\cdots,n-1}$ and $M\in\set{-L,-L+1,L}$ index a
total spin and its projection onto a quantization axis;
$I\equiv\p{n-1}/2$ is the maximal angular momentum of an $n$-level
spin; $\mu,\nu\in\set{-I,-I+1,\cdots,I}$ index projections of an
$n$-level nuclear spin onto a quantization axis; and
$\bk{I\mu;I\nu,LM}$ is a Clebsch-Gordan coefficient.  Similarly to the
GGM operators in \eqref{eq:GGM_diag} and \eqref{eq:GGM_off_diag}, the
transition operators in \eqref{eq:trans_ops} provide a complete basis
for the space of operators on the Hilbert space of an $n$-level
system, and satisfy the orthonormality condition
\begin{align}
  \obk{T_{LM}|T_{L'M'}} = \delta_{LL'} \delta_{MM'},
\end{align}
which implies
\begin{align}
  H_{\t{int}} = -\f12 u \sum_{p,q} {\v T^{(p)}}^\dag \c \v T^{(q)}
  = -\f12 u \v\T^\dag \c \v\T,
  &&
  \v\T \equiv \sum_p\v T^{(p)},
  \label{eq:H_int_trans}
\end{align}
where $\v T$ is a vector of all transition operators in
\eqref{eq:trans_ops}.

@@

Unlike the GGM operators in \eqref{eq:GGM_diag} and
\eqref{eq:GGM_off_diag}, the transition operators in
\eqref{eq:trans_ops} are not self-adjoint.  In order to express
Hamiltonians in terms of self-adjoint operators, we define the {\it
  drive operators}
\begin{align}
  D_{L,0} \equiv \sqrt{2}~ T_{L,0},
  &&
  D_{LM} \stackrel{M>0}{\equiv} -\p{T_{L\Delta} + T_{L\Delta}^\dag},
  &&
  D_{LM} \stackrel{M<0}{\equiv} i\p{T_{L\Delta} - T_{L\Delta}^\dag},
  \label{eq:drive_ops}
\end{align}
for integer $M$ with $\abs{M}\le L$.  Similarly to the GGM operators
in \eqref{eq:GGM_diag} and \eqref{eq:GGM_off_diag}, the drive
operators are equal to the Pauli operators in the case of $n=2$, and
satisfy the orthonormality condition
\begin{align}
  \obk{D_{LM}|D_{L'M'}} = 2\delta_{LL'}\delta_{MM'},
\end{align}
where $a,b$ index any drive operator in \eqref{eq:drive_ops}.  In
terms of the drive operators,
\begin{align}
  H_{\t{int}} = -\f14 u \sum_{p,q} \v D^{(p)}\c\v D^{(q)}
  = -u \v\D \c \v\D,
  &&
  \v\D \equiv \f12 \sum_p \v D^{(p)},
  \label{eq:H_int_drive}
\end{align}
where $\v D$ is a vector of all drive operators in
\eqref{eq:drive_ops}.

The drive operators have the convenient feature that $D_{1,0}$ and
$D_{1,1}$ are (up to a gauge transformation) respectively proportional
to the Hamiltonians induced on an $n$-level spin by an external
magnetic field and a classical driving laser; loosely speaking, one
should identify $\p{D_{1,0},D_{1,1}}\sim\p{S_\z,S_\x}$, where
$S_\alpha$ is a collective SU(2) spin-$\alpha$
operator\cite{perlin2019shorttime}.  More concretely, we can embed the
Hilbert space of a single $n$-level spin into the
permutationally-symmetric (Dicke) manifold of states for $\p{n-1}$
2-level spins via
\begin{align}
  \ket{\mu}_{\t{single}} \to \ket{\mu+I}_{\t{collective}}
  \propto S_+ \ket\dn^{\otimes\p{n-1}},
\end{align}
where $I\equiv\p{n-1}/2$ is a total spin,
$\mu\in\set{-I,-I+1,\cdots,I}$ indexes a state of the single $n$-level
spin, and $S_+$ is a collective spin-raising operator for $\p{n-1}$
2-level spins.  This embedding takes
\begin{align}
  \p{D_{1,0},D_{1,1},D_{1,-1}}
  \to \p{S_\z,S_\x,S_\y}
  \times \sqrt{\f{3}{I \p{I+1/2} \p{I+1}}}.
\end{align}

@@

Here we take the single-particle spin-orbit coupled Hamiltonian
$H_{\t{lat}}^{(\phi)}$ in \eqref{eq:H_lat_SOC}, and derive its
expression in \eqref{eq:H_lat_drive} in terms of the drive operators
$D_{LM}$.  In the basis of ordinary spin operators $s_{\mu\mu}$, this
Hamiltonian takes the form
\begin{align}
  H_{\t{lat}}^{(\phi)}
  = -J \sum_{q,\mu} \cos\p{qa+\mu\phi} s_{\mu\mu}^{(q)}
  \label{eq:H_lat_spin}
\end{align}
Defining the diagonal ``spin-wave'' operators
\begin{align}
  \omega_{\phi,+} \equiv \sum_\mu \cos\p{\mu\phi} s_{\mu\mu},
  &&
  \omega_{\phi,-} \equiv \sum_\mu \sin\p{\mu\phi} s_{\mu\mu},
\end{align}
which are respectively even and odd under spin inversion
($\mu\to-\mu$), we can simplify
\begin{align}
  \sum_\mu \cos\p{qa+\mu\phi} s_{\mu\mu}
  = \cos\p{qa} \omega_{\phi,+} - \sin\p{qa} \omega_{\phi,-},
\end{align}
and resolve the identity (super-)operator $\I$ with
$\I \omega_{\phi,\pm} = \omega_{\phi,\pm}$ in the basis of the drive
operators $D_{LM}$, finding
\begin{align}
  \sum_\mu \cos\p{qa+\mu\phi} s_{\mu\mu}
  = \cos\p{qa} \f12 \sum_{L,M} \obk{\omega_{\phi,+}|D_{LM}} D_{LM}
  - \sin\p{qa} \f12 \sum_{L,M} \obk{\omega_{\phi,-}|D_{LM}} D_{LM}.
\end{align}
We now note that the drive operators $D_{LM}$ are strictly
off-diagonal for $M\ne0$, and furthermore that $D_{L,0}$ with even
(odd) $L$ is even (odd) under spin inversion, which implies
\begin{align}
  \sum_{L,M} \obk{\omega_{\phi,+}|D_{LM}} D_{LM}
  = \sum_{L\,\t{even}} \obk{\omega_{\phi,+}|D_{L,0}} D_{L,0},
\end{align}
and likewise with $\omega_{\phi,-}$ and odd $L$.  In total, we can
write
\begin{align}
  \sum_\mu \cos\p{qa+\mu\phi} s_{\mu\mu}
  = \cos\p{qa} \f12 \sum_{L\,\t{even}}
  \obk{\omega_{\phi,+}|D_{L,0}} D_{L,0}
  - \sin\p{qa} \f12 \sum_{L\,\t{odd}}
  \obk{\omega_{\phi,-}|D_{L,0}} D_{L,0},
\end{align}
or, more compactly,
\begin{align}
  \sum_\mu \cos\p{qa+\mu\phi} s_{\mu\mu}
  = \sum_L B_L\p{q,\phi} D_{L,0},
  \label{eq:compact_SOC_drive}
\end{align}
with effective driving field strengths
\begin{align}
  B_L\p{q,\phi}
  &\equiv \p{-1}^L f_L\p{qa} \f12
  \sum_\mu f_L\p{\mu\phi} \obk{s_{\mu\mu}|D_{L,0}} \\
  &= \p{-1}^L f_L\p{qa} \sqrt{\f{L+1/2}{n}}
  \sum_\mu \bk{I\mu;I\mu,L,0} f_L\p{\mu\phi},
\end{align}
where
\begin{align}
  f_L\p{\theta} \equiv
  \begin{cases}
    \cos\theta & L~\t{even} \\
    \sin\theta & L~\t{odd}
  \end{cases}.
\end{align}
Substituting the result in \eqref{eq:compact_SOC_drive} into the
single-particle Hamiltonian in \eqref{eq:H_lat_spin} yields
\begin{align}
  H_{\t{lat}}^{(\phi)} = -J \sum_{q,L} B_L\p{q,\phi} D_{L,0}^{(q)}.
\end{align}
